---
title: 'Practical Machine Learning Course Project : Prediction Assignment'
author: "Cecilia Cruz-Ram, MD DPCOM"
date: "2/23/2020"
output: html_document
fontsize: 10pt
geometry: margin = 0.5in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions:

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

In this project, the goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 

The goal is to predict the manner in which the exercise was done.


## Simulations:

# A. Free Up Memory
```{r}
rm(list=ls())
```

# B. Setwd
```{r}
setwd("~/Documents/Files/Zetch/Online Courses/Data Science Specialization Feb18/R/Practical Machine Learning")
```

# C. Load Data
```{r}
library(caret)
library(knitr)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(corrplot)
library(gbm)
set.seed(12345)
```

```{r}
# Set the URL for the download
UrlTrain <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
UrlTest  <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

# Download Datasets
training <- read.csv(url(UrlTrain))
testing  <- read.csv(url(UrlTest))

# Create a partition with the training dataset 
inTrain  <- createDataPartition(training$classe, p=0.7, list=FALSE)
TrainSet <- training[inTrain, ]
TestSet  <- training[-inTrain, ]
dim(TrainSet)
dim(TestSet)
```
   
  Data sets consists of 5885 values with 160 variables.

# D. Clean and Pre-process Data

  Remove values with Near Zero Variance (NZV).
```{r}
NZV <- nearZeroVar(TrainSet)
TrainSet <- TrainSet[, -NZV]
TestSet  <- TestSet[, -NZV]
dim(TrainSet)
dim(TestSet)
```

  Remove values with NA.
```{r}
AllNA    <- sapply(TrainSet, function(x) mean(is.na(x))) > 0.95
TrainSet <- TrainSet[, AllNA==FALSE]
TestSet  <- TestSet[, AllNA==FALSE]
dim(TrainSet)
dim(TestSet)
```

   Remove identification only variables (columns 1 to 5)
```{r}
TrainSet <- TrainSet[, -(1:5)]
TestSet  <- TestSet[, -(1:5)]
dim(TrainSet)
dim(TestSet)
```

   The number of variables has been reduced to 54.

# D. Analyze Data

```{r}
corMatrix <- cor(TrainSet[, -54])
corrplot(corMatrix, order = "FPC", method = "circle", type = "lower", 
         tl.cex = 0.35, tl.col = rgb(0, 0, 0))
```

The highly correlated variables are shown in dark colors in the graph above. 

# E. Prediction Modelling

   Three methods will be applied to model the regressions (in the Train dataset) and the one with higher accuracy when applied to the Test dataset will be used for the quiz predictions. The methods are: Random Forests, Decision Tree and Generalized Boosted Model.
   
   A Confusion Matrix is plotted at the end of each analysis to better visualize the accuracy of the models.
   
   1) Random Forest
```{r}
# Model fit
set.seed(12345)
controlRF <- trainControl(method = "cv", number = 3, verboseIter = FALSE)
modFitRandForest <- train(classe ~ ., data = TrainSet, method = "rf",
                          trControl = controlRF)
modFitRandForest$finalModel
```

```{r}
# Prediction on Test Dataset
predictRandForest <- predict(modFitRandForest, newdata = TestSet)
confMatRandForest <- confusionMatrix(predictRandForest, TestSet$classe)
confMatRandForest
```

```{r}
# Plot Matrix Results
plot(confMatRandForest$table, col = confMatRandForest$byClass, 
     main = paste("Random Forest Accuracy =",
                  round(confMatRandForest$overall['Accuracy'], 3)))
```

 2) Decision Tree
```{r}
# Model fit
set.seed(12345)
modFitDecTree <- rpart(classe ~ ., data = TrainSet, method = "class")
fancyRpartPlot(modFitDecTree)
```

```{r}
# Prediction on Test Dataset
predictDecTree <- predict(modFitDecTree, newdata = TestSet, type = "class")
confMatDecTree <- confusionMatrix(predictDecTree, TestSet$classe)
confMatDecTree
```

```{r}
# Plot Matrix Results
plot(confMatDecTree$table, col = confMatDecTree$byClass, 
     main = paste("Decision Tree Accuracy =",
                  round(confMatDecTree$overall['Accuracy'], 3)))
```

   3) Generalized Boosted Model
```{r}
# Model fit
set.seed(12345)
controlGBM <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
modFitGBM  <- train(classe ~ ., data = TrainSet, method = "gbm",
                    trControl = controlGBM, verbose = FALSE)
modFitGBM$finalModel
```

```{r}
# Prediction on Test Dataset
predictGBM <- predict(modFitGBM, newdata=TestSet)
confMatGBM <- confusionMatrix(predictGBM, TestSet$classe)
confMatGBM
```

```{r}
# Plot Matrix Results
plot(confMatGBM$table, col = confMatGBM$byClass, 
     main = paste("GBM Accuracy =", round(confMatGBM$overall['Accuracy'], 3)))
```

# F. Model Application to Test Data

   The accuracy of the 3 regression modelling methods are:

      Random Forest : 0.999
      Decision Tree : 0.734
      GBM           : 0.987

   The Random Forest model will be applied to predict the 20 quiz results (testing dataset).
   
```{r}
predictTEST <- predict(modFitRandForest, newdata=testing)
predictTEST
```

____________________________________

## References

1. Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.

2. http://groupware.les.inf.puc-rio.br/har, visited 2015/12/25